---
title: "Lecture 11: Gaussian Process"
author: "Chun-Hao Yang"
footer: "[Home](https://chunhaoy.com/stat-5010/)"
format: 
    revealjs:
        theme: slides.scss
        chalkboard: true
        slide-number: true
        html-math-method: mathjax
        incremental: true
        scrollable: true
        include-in-header: 
            - text: 
                <link href='https://fonts.googleapis.com/css?family=Fira Code' rel='stylesheet'>
execute:
    echo: false
    warning: false
    cache: true
    freeze: true
---

## Non-, Semi-, and Parametric Models

{{< include macro.qmd >}}

-   What we have learned so far are called **parametric models**, i.e., models that are determined by a finite number of parameters.
-   **Nonparametric models** are models whose ``parameter'' is infinite dimensional.
-   Two common types of nonparametric models:
    +   **Nonparametric regression**: $Y = f(X) + \epsilon$, where $f$ is an unknown function.
    +   **Nonparametric density estimation**: $X_1, \ldots, X_n \iid F$, where $F$ is an unknown distribution (without assuming any parametric family).
-   **Semiparametric models** are models that have both parametric and nonparametric components, e.g., $Y = X^T\beta + f(Z) + \epsilon$.

## Bayesian Nonparametrics

-   For a nonparametric model, a Bayesian treats the unknown function/distribution as the parameter.
-   Therefore, we need prior distributions for these parameters.
-   We will introduce two "distributions" that are useful for nonparametric models:
    +   Gaussian process (GP, the "distribution" of random functions)
    +   Dirichlet process (DP, the "distribution" of random distributions)
-   That is,
    +   $Y = f(x) + \epsilon$, $f \sim \text{GP}$, $\epsilon \sim N(0, \sigma^2)$.
    +   $X_1, \ldots, X_n \sim F$, $F \sim \text{DP}$.
    
## Why nonparametrics?

-   Nonparametric models are more flexible than parametric models.
-   Yet, nonparametric models are not always better than parametric models.
-   As we increase the complexity of the model, we may 
    +   overfit the data;
    +   increase the computational cost;
    +   lose interpretability.
-   ??

## Nonparametric regression[^cmu_note]

-   Recall that a linear regression model assumes $\E(Y \mid X) = X^T\beta$.
-   A nonparametric regression model assumes $\E(Y \mid X) = f(X)$, where $f$ is an unknown function.
-   There are different ways to estimate $f$:
    +   Kernel regression
    +   Regression splines
    +   Gaussian process regression
-   For simplicity, we consider the one-dimensional regression problem, i.e., we have  $\{x_i, y_i\}_{i=1}^n$ where $x_i, y_i \in \R$.
-   The extension to higher dimensions is straightforward.


[^cmu_note]: A good reference is this [course note](https://www.stat.cmu.edu/~larry/=sml/nonpar.pdf).
    
## Kernel regression

-   The Nadaraya–Watson kernel regression estimator of $f$ is 
$$
\hat{f}(x) = \frac{\sum_{i=1}^n K_h(x - X_i) Y_i}{\sum_{i=1}^n K_h(x - X_i)},
$$
where $K_h(x) = K(x/h)/h$ is a kernel function, and $h$ is a bandwidth parameter.
-   That is, $\hat{f}(x)$ is a weighted average of the $Y_i$'s, where the weights are determined by the kernel function $K$.
-   The kernel function $K$ is usually chosen to be a symmetric density function, e.g., the standard normal density function.

## Example

```{r, fig.height=6, fig.width=8}
#| label: kernel_regression
#| echo: true
#| fig-align: center
#| output-location: slide
library(gplm) # for kreg()
set.seed(1)
n <- 1000
x <- rnorm(n)
y <- sin(x) + rnorm(n)
plot(x,y,col="gray")
curve(sin(x), add = T, col = "green")
lines(kreg(x,y), lwd = 2, col = "blue")
```



## Regression splines

-   A function $g$ is called a [$k$th order spline with knots $t_0 < \cdots < t_m$]{.underline} if
    +   $g$ is a polynomial of degree $k$ on the intervals $(-\infty, t_0], [t_0, t_1],\ldots, [t_m, \infty)$.
    +   $g$ has continuous derivatives up to order $k-1$ at each knot $t_i$.
-   Given a set of knots $t_0 < \cdots < t_m$, the space of $k$th order spline has dimension $m + k + 1$.
-   There are many choices of basis for the space of splines. The most popular one is the **B-spline** basis.
-   Hence a regression spline model is 
    $$
    \E(Y \mid x) = \sum_{j=1}^{m + k + 1} \beta_j B_j(x),
    $$ 
    where $B_j$ is the $j$th B-spline basis function.
    
## Regression splines

-   Although a regression spline is determined by $m+k+1$ parameters, it is usually referred to as a nonparametric model, since we are not interested in those parameters.
-   The choice of knots is important. 
-   A **smoothing spline** uses all inputs as knots and avoids overfitting by shrinking the coefficients.
-   A **natural spline** assumes polynomial of degree $(k-1)/2$ on $(-\infty, t_0]$ and $[t_m, \infty)$ to reduce variance at the boundary.
-   Natural spline with cubic order is the most common special case.

## Example

```{r, fig.height=6, fig.width=8}
#| label: spline
#| fig-align: center
library(gplm) # for kreg()
set.seed(1)
n <- 1000
x <- rnorm(n)
y <- sin(x) + rnorm(n)
plot(x,y,col="gray")
curve(sin(x), add = T, col = "green")
lines(kreg(x,y), lwd = 2, col = "blue")
lines(smooth.spline(x,y), lwd = 2, col = "red")
```


# Gaussian Process

## Definition

A ``random function'' $f$ is said to follow a Gaussian process, denoted by 
$$
f \sim \mc{GP}(\mu, K),
$$
if for **any** $x_1, \ldots, x_n$, the random vector $(f(x_1), \ldots, f(x_n))$ has a multivariate normal distribution, i.e.,
$$
\left[\begin{array}{c}
f(x_1)\\
\vdots\\
f(x_n)
\end{array}\right] \sim N\left(\left[\begin{array}{c}
\mu(x_1)\\
\vdots\\
\mu(x_n),
\end{array}\right], \left[\begin{array}{ccc}
K(x_1, x_1) & \cdots & K(x_1, x_n)\\
\vdots & \ddots & \vdots\\
K(x_n, x_1) & \cdots & K(x_n, x_n)
\end{array}\right]\right).
$$

-   The parameter $\mu: \R \to \R$ is called the **mean function**.
-   The parameter $K: \R \times \R \to \R$ is called the **covariance function/operator** or **kernel**.
-   The kernel $K$ needs to be symmetric and positive definite, i.e, for any $x_1, \ldots, x_n \in \R$, the matrix above is symmetric and positive definite.

## Commonly used kernels

-   Notation: for $x, x^{\prime} \in \R^n$, $K(x, x^{\prime})$ is an $n \times n$ matrix whose $(i,j)$th entry is $K(x_i, x^{\prime}_j)$.

-   Linear kernel: $K(x, x^{\prime}) = x^Tx^{\prime}$.
-   Polynomial kernel: $K_{c,d}(x, x^{\prime}) = (x^Tx^{\prime} + c)^d$.
-   Gaussian kernel: $K_{\sigma}(x, x^{\prime}) = \exp\left(-\frac{\|x-x^{\prime}\|^2}{2\sigma^2}\right)$.

## Simulating Gaussian processes

```{r, fig.width = 10}
#| label: GP_sim
#| echo: true
library(mvtnorm)
GP_sim <- function(from = 0, to = 1, mean_func = function(x){0},
                   cov_func = function(x1, x2){exp(-16*(x1-x2)^2)}, 
                   m = 500){
    x <- seq(from, to, length.out = m)
    mu <- sapply(x, mean_func)
    Sigma <- outer(x, x, Vectorize(cov_func))
    y <- rmvnorm(1, mu, Sigma)
    return(list(x = x, y = y))
}
```

## Simulating Gaussian processes

```{r, fig.width = 10}
#| label: GP_sim_plot
#| fig-align: center
library(latex2exp)
set.seed(2023)
par(mfrow = c(1, 2))
plot(0, type = 'n', xlim = c(0, 1), ylim = c(-3, 3), 
     xlab = 'x', ylab = 'f(x)',
     main = TeX('$\\mu(x) = 0$, Gaussian kernel'))
abline(h = 0, lty = 2, lwd = 2)
for(i in 1:10){
    f <- GP_sim()
    lines(f$x, f$y, col = i)
}
plot(0, type = 'n', xlim = c(0, 1), ylim = c(-3, 3), 
     xlab = 'x', ylab = 'y',
     main = TeX('$\\mu(x) = 0$, $K(x_1, x_2) = \\min(x_1, x_2)$'))
abline(h = 0, lty = 2, lwd = 2)
for(i in 1:10){
    f <- GP_sim(cov_func = function(x1, x2){min(c(x1, x2))})
    lines(f$x, f$y, col = i)
}
```

## Gaussian Process Regression

-   Suppose we observe $\{x_i, y_i\}_{i=1}^n$, $x_i, y_i \in \R$.
-   Let $\mathbf{y} = [y_1, \ldots, y_n]^T$ and $\mathbf{x} = [x_1, \ldots, x_n]^T$.
-   Consider the model:
    \begin{align*}
    \mathbf{Y} \mid f, \mathbf{x}, \sigma^2 &\sim N(f(\mathbf{x}), \sigma^2I_n)\\
    f \mid \mu, K &\sim \mc{GP}(\mu, K).
    \end{align*}
-   The GP prior is equivalent to $f(\mathbf{x}) \mid \mathbf{x}, \mu, K \sim N(\mu(\mathbf{x}), K(\mathbf{x}, \mathbf{x}))$.
-   The likelihood is equivalent to $\mathbf{Y} = f(\mathbf{x}) + \epsilon$, $\epsilon \sim N(0, \sigma^2I_n)$.
-   We need to compute two distributions:
    +   The posterior distribution of $f$ given $\mathbf{y}$ and $\mathbf{x}$.
    +   The posterior predictive distribution of $f(\mathbf{x}^{\prime})$ given $\mathbf{y}$ and $\mathbf{x}$.

    
    
## Posterior Predictive Distribution
-   The posterior distribution of $f$ is also a Gaussian process
    $$
    f \mid \mathbf{y}, \mathbf{x}, \sigma^2 \sim \mc{GP}(\bar{\mu}, \bar{K})
    $$
    where
    \begin{align*}
    \bar{\mu}(\cdot) & = \mu(\cdot) + K(\cdot, \mathbf{x})[K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n]^{-1}(\mathbf{y} - \mu(\mathbf{x}))\\
    \bar{K}(\cdot, \cdot) & = K(\cdot, \cdot) - K(\cdot, \mathbf{x})[K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n]^{-1}K(\mathbf{x}, \cdot).
    \end{align*}
-   Hence the predictive distribution is a multivariate normal
    $$
    f(\mathbf{x}^{\prime}) \mid \mathbf{y}, \mathbf{x}, \sigma^2 \sim N(\bar{\mu}(\mathbf{x}^{\prime}), \bar{K}(\mathbf{x}^{\prime}, \mathbf{x}^{\prime})).
    $$

## Derivation

-   For any $m$ and $\mathbf{x}^{\prime} \in \R^m$,
    $$
    \left[\begin{array}{c}
    \mathbf{Y}\\
    f(\mathbf{x}^{\prime})
    \end{array}
    \right] \sim N\left(\left[\begin{array}{c}
    \mu(\mathbf{x})\\
    \mu(\mathbf{x}^{\prime})
    \end{array}\right], \left[\begin{array}{cc}
    K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n& K(\mathbf{x}, \mathbf{x}^{\prime})\\
    K(\mathbf{x}^{\prime}, \mathbf{x}) & K(\mathbf{x}^{\prime}, \mathbf{x}^{\prime})
    \end{array}\right]\right).
    $$
-   Therefore
    $$
    f(\mathbf{x}^{\prime}) \mid \mathbf{y}, \mathbf{x}, \sigma^2 \sim N\left(\bar{\mu}(\mathbf{x}^{\prime}) , \bar{K}(\mathbf{x}^{\prime}, \mathbf{x}^{\prime})\right).
    $$
    where
    \begin{align*}
    \bar{\mu}(\mathbf{x}^{\prime}) & = \mu(\mathbf{x}^{\prime}) + K(\mathbf{x}^{\prime}, \mathbf{x})[K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n]^{-1}(\mathbf{y} - \mu(\mathbf{x}))\\
    \bar{K}(\mathbf{x}^{\prime}, \mathbf{x}^{\prime}) & = K(\mathbf{x}^{\prime}, \mathbf{x}^{\prime}) - K(\mathbf{x}^{\prime}, \mathbf{x})[K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n]^{-1}K(\mathbf{x}, \mathbf{x}^{\prime}).
    \end{align*}
-   That is, $f \mid \mathbf{y}, \mathbf{x}, \sigma^2 \sim \mc{GP}(\bar{\mu}, \bar{K})$


## Example

Take $n = 20$ and generate $Y = \sin(X) + \exp(X/5) + \epsilon$ where $\epsilon \sim N(0, 0.1^2)$.


```{r, fig.height=4, fig.width=6}
#| label: GP_example_data
#| fig.align: center
n <- 20
set.seed(123)
X <- runif(n, -5, 5)
Y <- sin(X) + exp(X/5) + rnorm(n, 0, 0.1)
plot(X, Y, pch = 16, cex = 1.5)
```

## Example

Fit a Gaussian process regression model with $\mu(x) = 0$ and Gaussian Kernel.

```{r, fig.height=4, fig.width=6}
#| label: GP_example_fit
#| echo: true
#| fig.align: center
library(GauPro)
gp <- GauPro(X, Y)
plot(gp)
curve(sin(x) + exp(x/5), -5, 5, add = TRUE, col = "red", lwd = 2)
```




## Computational issues

-   It seems that posterior computation in Gaussian process regression is trivial, but there are two main hurdles involved.
-   The first hurdle is that the covariance matrix $K(\mathbf{x}, \mathbf{x}) + \sigma^2I_n$ is $n \times n$ and the computational complexity of the inverse matrix is $O(n^3)$.
-   The second hurdle is that the posterior distribution is high dimensional, making the sampling inefficient.


-   We can rewrite the model as 
    \begin{align*}
    Y \mid f, \sigma^2 &\ind N(f, \sigma^2I_n)\\
    f \mid x &\sim N(\mu, K(x, x))\\
    \sigma^2 &\sim \text{IG}(a, b)
    \end{align*}
-   Suppose $a$ and $b$ are known hyperparameters.
-   The posterior distribution is $\pi(f, \sigma^2 \mid \mathbf{y}, \mathbf{x})$, which is in $(n+1)$-dimension.




## Example 

```{stan output.var = "GP_model"}
#| echo: true
data {
  int<lower=1> N;
  real x[N];
}
transformed data {
  matrix[N, N] K = cov_exp_quad(x, 1.0, 1.0); // l = 1.0, tau^2 = 1.0
  vector[N] mu = rep_vector(0, N); // mu = 0
  for (n in 1:N)
    K[n, n] = K[n, n] + 0.1; // sigma^2 = 0.1
}
parameters {
  vector[N] y;
}
model {
  y ~ multi_normal(mu, K);
}
```

## Example

```{r}
#| echo: true
library(rstan)
data <- list(N = 100, x = seq(0, 1, length.out = 100))
GP_sample <- sampling(GP_model, data = data, chains = 1, iter = 1000,
                      refresh = 0)

y <- GP_sample@sim$samples[[1]][1:100] |>
    unlist() |>
    matrix(ncol = 100, byrow = TRUE)

plot(0, xlim = c(0, 1), ylim = c(-3, 3), xlab = "x", ylab = "y")
for(i in 1:10){
  lines(data$x, y[i,], col = "grey", lwd = 0.5)
}


```

## Hierarchical Gaussian Process model

-   Consider a fully Bayesian hierarchical model
    \begin{align*}
    \rho & \sim \text{InvGamma}(5,5) \\
    \alpha & \sim N(0,1) \\
    \sigma & \sim N(0,1) \\
    f_i & \iid N(0, K(x \mid \alpha, \rho)) \\
    y_i & \ind N\left(f_i, \sigma^2\right) \quad i = 1, \ldots, n.
    \end{align*}

## Marginal likelihood GP


## Latent Gaussian Process model


